---
title: "RMS Titanic"
author: "Maya Cranor"
date: 2025
output:
  github_document:
    toc: true
prerequisites:
  - e-vis01-bar-charts
---

*Purpose*: Most datasets have at least a few variables. Part of our task in analyzing a dataset is to understand trends as they vary across these different variables. Unless we're careful and thorough, we can easily miss these patterns. In this challenge you'll analyze a dataset with a small number of categorical variables and try to find differences among the groups.

*Reading*: (Optional) [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|----|----|----|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**

```{r setup}
library(tidyverse)

df_titanic <- as_tibble(Titanic)
```

*Background*: The RMS Titanic sank on its maiden voyage in 1912; about 67% of its passengers died.

# First Look

<!-- -------------------------------------------------- -->

### **q1** Perform a glimpse of `df_titanic`. What variables are in this dataset?

```{r q1-task}
## TASK: Perform a `glimpse` of df_titanic
df_titanic %>% 
  glimpse()
```

**Observations**:

-   Class: 1st, 2nd, 3rd, Crew
-   Sex: Male, Female
-   Age: Child, and Adult
-   Survived: Yes, No
-   n: number of people

### **q2** Skim the [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic, and look for a total count of souls aboard. Compare against the total computed below. Are there any differences? Are those differences large or small? What might account for those differences?

```{r q2-summary}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_titanic %>% summarize(total = sum(n))
```

**Observations**:

-   Write your observations here
    -   Wikipedia article states that there were an estimated 2,224 passengers and crew, but the data set has a total of 2201 people.
-   Are there any differences?
    -   There are 23 people less in the dataset than wikipedia
-   If yes, what might account for those differences?
    -   People deciding to not go last minute, might not have reported that to the ship. Additionally there were people that used alias that could have been double counted. It would have also been harder to keep track of people in 1912 since records were kept on paper.
    -   ...

### **q3** Create a plot showing the count of persons who *did* survive, along with aesthetics for `Class` and `Sex`. Document your observations below.

*Note*: There are many ways to do this.

```{r q3-task}
## TASK: Visualize counts against `Class` and `Sex`
df_titanic %>% 
  filter(Survived == "Yes") %>% 
  ggplot(aes(x = Class, y = n, fill = Sex)) +
  geom_col(position = "dodge")
```

**Observations**:

-   Crew was the largest group of people to survive closely followed by 1st class and then by 3rd class. Crew is the biggest class group so it makes sense that it had the most survivors. 1st class had the most prioritization as it had about the same amount of survivors as crew but with half the original size. Since the 1st is paying the most/has the most celebrity it could lead to more of them.
-   Female sex was also heavily prioritized for saving as each group had a higher percentage of females saved versus there original group size. Crew was the only group that still had a strong majority of surviving males, but they had very little females to begin with.

# Deeper Look

<!-- -------------------------------------------------- -->

Raw counts give us a sense of totals, but they are not as useful for understanding differences between groups. This is because the differences we see in counts could be due to either the relative size of the group OR differences in outcomes for those groups. To make comparisons between groups, we should also consider *proportions*.[1]

The following code computes proportions within each `Class, Sex, Age` group.

```{r compute-prop}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_prop <-
  df_titanic %>%
  group_by(Class, Sex, Age) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup()
df_prop
```

### **q4** Replicate your visual from q3, but display `Prop` in place of `n`. Document your observations, and note any new/different observations you make in comparison with q3. Is there anything *fishy* in your plot?

```{r q4-task}

df_prop %>% 
  filter(Survived == "Yes") %>% 
  ggplot(aes(x = Class, y = Prop, fill = Sex)) +
  geom_col(position = "dodge")
```

**Observations**:

-   Write your observations here.
    -   The graph shows that 100% percent of 1st and 2nd class male and female people survived. However less than half of 3rd class males and 3rd class females survived. Most female crew remembers survived, and less of than a quarter of the male crew members survived. For all classes either the same amount or more females survived than there male counter parts.
-   Is there anything *fishy* going on in your plot?
    -   From looking at the data in a table format, I know that not all 1st and 2nd class members survived. The value of prop should be less than 1 for those columns. I believe that age is not being represented in the data. Since I am using position = "dodge" information with the same class and gender labels are getting overlaid on top of each other. So between adult and child for each category, which ever one has a higher prop will get obviously displayed.

### **q5** Create a plot showing the group-proportion of occupants who *did* survive, along with aesthetics for `Class`, `Sex`, *and* `Age`. Document your observations below.

*Hint*: Don't forget that you can use `facet_grid` to help consider additional variables!

```{r q5-task}
df_prop %>% 
  filter(Survived == "Yes") %>% 
  ggplot(aes(x = Class, y = Prop, fill = Sex)) +
  geom_col(position = "dodge")+
  facet_wrap(~Age)
```

**Observations**:

-   For adults, a majority of women survived compared to their male counter parts. 1st class also had the highest proportion for each of the sexes compared to the other classes. However proportionally woman across all classes for adults were higher than the highest proportion of men. All children that are in the 1st and 2nd class survived, and a bit less than half of the children in 3rd class survived. Part of that could be explained by there being only 30 children total between 1st and 2nd class and 79 children total for 3rd class, so there were more overall children that had to be saved in 3rd class. From this it appears that woman and children survived more than adult males.\
-   If you saw something *fishy* in q4 above, use your new plot to explain the fishy-ness.
    -   This graph separates out age from the q4 graph. Now adult and children data aren't getting overlaid on top of each other.

# Notes

<!-- -------------------------------------------------- -->

[1] This is basically the same idea as [Dimensional Analysis](https://en.wikipedia.org/wiki/Dimensional_analysis); computing proportions is akin to non-dimensionalizing a quantity.
