---
title: "Gapminder"
author: "Maya Cranor"
date: 2025-
output:
  github_document:
    toc: true
prerequisites:
  - e-vis03-boxplots
---

*Purpose*: Learning to do EDA well takes practice! In this challenge you'll further practice EDA by first completing a guided exploration, then by conducting your own investigation. This challenge will also give you a chance to use the wide variety of visual tools we've been learning.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|------------------|-----------------------------|-------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**

```{r setup}
library(tidyverse)
library(gapminder)

```

*Background*: [Gapminder](https://www.gapminder.org/about-gapminder/) is an independent organization that seeks to educate people about the state of the world. They seek to counteract the worldview constructed by a hype-driven media cycle, and promote a "fact-based worldview" by focusing on data. The dataset we'll study in this challenge is from Gapminder.

# Guided EDA

<!-- -------------------------------------------------- -->

First, we'll go through a round of *guided EDA*. Try to pay attention to the high-level process we're going through---after this guided round you'll be responsible for doing another cycle of EDA on your own!

### **q0** Perform your "first checks" on the dataset. What variables are in this

dataset?

```{r q0-task}
## TASK: Do your "first checks" here!
glimpse(gapminder)
head(gapminder)
```

**Observations**:

-   Write all variable names here
-   country: name of country on Earth
-   continent: name of continent on Earth
-   year: year
-   lifeExp: average life expediency in years
-   pop: population of the humans in the country
-   gdpPercap: gross domestic product - a country’s economic output to a per-person outlook

### **q1** Determine the most and least recent years in the `gapminder` dataset.

*Hint*: Use the `pull()` function to get a vector out of a tibble. (Rather than the `$` notation of base R.)

```{r q1-task}
## TASK: Find the largest and smallest values of `year` in `gapminder`
year_max <- gapminder %>% pull(year) %>% max()
year_min <- gapminder %>% pull(year) %>% min()

year_max
year_min
```

Use the following test to check your work.

```{r q1-tests}
## NOTE: No need to change this
assertthat::assert_that(year_max %% 7 == 5)
assertthat::assert_that(year_max %% 3 == 0)
assertthat::assert_that(year_min %% 7 == 6)
assertthat::assert_that(year_min %% 3 == 2)

if (is_tibble(year_max)) {
  print("year_max is a tibble; try using `pull()` to get a vector")
  assertthat::assert_that(False)
}

print("Nice!")
```

### **q2** Filter on years matching `year_min`, and make a plot of the GDP per capita against continent. Choose an appropriate `geom_` to visualize the data. What observations can you make?

You may encounter difficulties in visualizing these data; if so document your challenges and attempt to produce the most informative visual you can.

```{r q2-task}
## TASK: Create a visual of gdpPercap vs continent
gapminder %>% 
  filter(year == 1952) %>% 
  ggplot(aes(x = continent , y = gdpPercap)) +
  geom_jitter() 

## removed the outlier to get a better look at the data
gapminder %>% 
  filter(year == 1952, gdpPercap < 90000) %>% 
  ggplot(aes(x = continent , y = gdpPercap)) +
  geom_jitter() 

gapminder %>% 
  filter(year == 1952,  gdpPercap < 90000) %>% 
  ggplot(aes(x = continent , y = gdpPercap)) +
 geom_boxplot()
```

**Observations**:

-   Africa has the lowest average gdpPercap without any breakaway outliers.
-   Americas and Europe have the highest minimum gdpPercap, as their countries values are atleast above 2000.
-   Americas has a clumping of countries between 2000-5000 gdpPercap, and a handful of countries with higher gdpPercap. Two of those countries have much higher gdpPercaps as they afe both above 10000
-   Europe's countries are evenly distrubted over a range of roughly 2000 to 10000, with one outlier at 15000
-   Oceania has the smallest amount of countries at 2, but they both have a fairly similar gdpPercap at roughly 10000.
-   Asia has the widest range of values going from 700 to over 100,000 gdpPercap. Most countries are near the 1000 gdpPercap, while there is one extreme outlier (Kuwait), which at least 7x bigger than other country.

**Difficulties & Approaches**:

-   Write your challenges and your approach to solving them
-   Asia has an extreme outlier, which changes the scale of the graph greatly, as it at least 7x bigger than every other value. To get a better look at the trends happening in the other countries, I filtered out the outlier. Once I did that, I was able to see what was happening!
-   Another issue I had was that it was hard to see all the values plotted as a scatter plot since they were stacked on top of each other. To solve this, I used the jitter feature, which slightly shifts the points. Since the X-axis only had the continent, as long as the dot was still grouped together with its other continent dots, it didn't need to be precisely on the line. This allowed me to see how many dots there were with similar values.

### **q3** You should have found *at least* three outliers in q2 (but possibly many more!). Identify those outliers (figure out which countries they are).

```{r q3-task}
## TASK: Identify the outliers from q2
#highlight outliers?

gapminder %>%
  filter(year == 1952) %>%
  summarise(mean_gdp = mean(gdpPercap, na.rm = TRUE), median_gdp = median(gdpPercap, na.rm = TRUE))


gapminder %>% 
  glimpse() %>% 
  filter(year == 1952, gdpPercap > 13990)


# gapminder %>%
#   glimpse() %>%
#   filter(year == 2007, gdpPercap > 37000)


```

**Observations**:

-   Identify the outlier countries from q2
    -   Outliers: Kuwait, Switzerland, United States
    -   I picked these countries as outliers because they are the 3 countries with highest gpdPercap. The mean gdpPercap of all countries is 3725.276 and the median is 1968.528, and Kuwait, Switzerland, United States gdpPercap are all above 13,000. Kuwait is an extreme outlier because it's gpdPercap is 108382.35 which is more 7 times bigger than Switzerland's which is the next biggest country.

*Hint*: For the next task, it's helpful to know a ggplot trick we'll learn in an upcoming exercise: You can use the `data` argument inside any `geom_*` to modify the data that will be plotted *by that geom only*. For instance, you can use this trick to filter a set of points to label:

```{r layer-filter}
## NOTE: No need to edit, use ideas from this in q4 below
gapminder %>%
  filter(year == max(year)) %>%

  ggplot(aes(continent, lifeExp)) +
  geom_boxplot() +
  geom_point(
    data = . %>% filter(country %in% c("United Kingdom", "Japan", "Zambia")),
    mapping = aes(color = country),
    size = 2
  )
```

### **q4** Create a plot similar to yours from q2 studying both `year_min` and `year_max`. Find a way to highlight the outliers from q3 on your plot *in a way that lets you identify which country is which*. Compare the patterns between `year_min` and `year_max`.

*Hint*: We've learned a lot of different ways to show multiple variables; think about using different aesthetics or facets.

```{r q4-task}
## TASK: Create a visual of gdpPercap vs continent
# For the year 1952
gapminder %>% 
  filter(year == 1952 | year == 2007) %>% 
  ggplot(aes(x = continent , y = gdpPercap)) +
  geom_jitter(
    data = . %>% 
      mutate(
        country = case_when(
          country %in% c("Kuwait", "Switzerland", "United States") ~ country,
          TRUE ~ "(other)"
        )
      ),
    mapping = aes(color = country)
  ) +
  scale_color_manual(
    values = c(
      "(other)" = "black",
      "Kuwait" = "red",
      "Switzerland" = "blue",
      "United States" = "green"
    )
  ) +
  facet_wrap(~ year)


```

**Observations**:

-   Overall lots of countries have increased their gdpPercap, likely due to inflation.
-   Kuwait has dramatically decreased its gpdPercap from 108382.35 to 47306.99, and is no longer an extreme outlier as it no longer has the highest gpdPercap and there are several countries with similar values.
-   US remains country with the highest gpdPercap in the Americas. It's gpdPercap has increased from 13990.48 to 42951.65.
-   Switzerland has the 2nd highest gpdPercap of all the countries or the highest of Europe. However, it still has a relatively high gpdPercap as its not the 7th highest country. It's gpdPercap has increased from 14734.23 to 37506.42

# Your Own EDA

<!-- -------------------------------------------------- -->

Now it's your turn! We just went through guided EDA considering the GDP per capita at two time points. You can continue looking at outliers, consider different years, repeat the exercise with `lifeExp`, consider the relationship between variables, or something else entirely.

### **q5** Create *at least* three new figures below. With each figure, try to pose new questions about the data.

```{r q5-task1}
## TASK: Your first graph
gapminder %>% 
  filter(year == 1952 | year == 2007 ) %>% 
  ggplot(aes(x = continent , y = lifeExp)) +
  geom_boxplot() +
  facet_wrap(~ year)
```

-   Graph shows the life expediencies in 1952 and 2007 of all countries.
-   The life expediency in every continent generally has increased.
-   The median order of life expediency has retained the same order relative to all of the continents, with Africa having the lowest life expediency and Oceania having the highest.
-   Asia had the biggest increase in life expediency, as it's median in 1952 was around 45 and in 2007 it was around 73. It also much closer to Americas and Europe's median life expediency, instead of Africa's.

```{r q5-task2}
## TASK: Your second graph
gapminder %>% 
  filter(year == 1952 | year == 2007) %>% 
  ggplot(aes(x = continent , y = pop)) +
  geom_jitter(
    data = . %>% 
      mutate(
        country = case_when(
          country %in% c("China", "India", "United States") ~ country,
          TRUE ~ "(other)"
        )
      ),
    mapping = aes(color = country)
  ) +
  geom_hline(
    data = . %>% group_by(year) %>% summarise(avg_pop = mean(pop, na.rm = TRUE)),
    aes(yintercept = avg_pop, color = "Average Population"),
    linetype = "dashed",
    size = 1
  ) +
  geom_hline(
    data = . %>% group_by(year) %>% summarise(median_pop = median(pop, na.rm = TRUE)),
    aes(yintercept = median_pop, color = "Median Population"),
    linetype = "dotted",
    size = 1
  ) +
  scale_color_manual(
    values = c(
      "(other)" = "black",
      "China" = "red",
      "India" = "blue",
      "United States" = "green",
      "Average Population" = "purple",
      "Median Population" = "orange"
    )
  ) +
  facet_wrap(~ year) +
  labs(color = "Country/Line")



```

-   Graph shows the population size in 1952 and 2007 of all countries.
-   Overall there seems to be an increase in population of all countries as the average and median population both increased.
-   China, India, United States have the biggest populations (in that order) in both 1952 and 2007. They also increased greatly from 1952 to 2007.

```{r q5-task3}
## TASK: Your third graph
gapminder %>% 
  filter(country == "Kuwait" ) %>% 
  ggplot(aes(x = year , y = gdpPercap)) +
  geom_jitter() 
```

-   Graph shows the gpdPercap in Kuwait from 1950 to 2007
-   After 1960 there is a decrease in gdpPercap, which continues to plummit until 1990s
    -   Either the population or general wealth changes drastically over this time period
